---
title: "Justin + Alden"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Loading the Data

```{r, include = FALSE}
set.seed(230)
library(car)
library(dplyr)
library(ggplot2)
library(readr)
```

```{r, include = FALSE}
#read in data
world_bank_2016 <- read_csv("Data/WB.2016.csv")

# remove NA data for measles and GNI
# wbCor <- wb[!is.na(wb$Measles),]
# wbCor <- wbCor[!is.na(wbCor$GNI),]
```

## Correlation Analysis

First, we would like to see whether global health indicators are associated with economic indicators such as GNI or percent of the population in rural areas.

We acknowledge that environment indicators such as PM2.5, ...., <fill in> may cause a degree of omitted variable bias, but we believe that ... <insert reason why only looking at health indicators.

```{r}
health_indicators <- world_bank_2016 %>%
  select(GNI, Rural, Measles, InfMort, LifeExp) %>%
  mutate(logGNI = log10(GNI)) %>%
  na.omit()
```

First, to test whether our hypothesis that GNI and Rural are good indicators of economic development....

TODO: Fix plot labels, etc.

```{r}
ggplot(data = health_indicators, aes(x = GNI, y = Rural)) +
  geom_point() +
  ggtitle("% Rural Population vs. GNI")
```

We see that they appear to be correlated, but not in a linear relationship. It appears that most countries have GNIs between 0 and 20000, which may violate a homoscedasticity assumption when running a correlation. Thus, in an effort to get to a more linear form, we transform GNI to logGNI.

```{r}
ggplot(data = health_indicators, aes(x = logGNI, y = Rural)) +
  geom_point() +
  ggtitle("% Rural Population vs. logGNI")
```

This looks much more linear! TODO: Check the other linear assumptions.

### 1. Hypotheses

TODO: In the hypothesis, should this be GNI or logGNI?

Null Hypothesis: There is no correlation between the percentage of a country living in a rural area and the GNI of a country. That is, the correlation equals 0.

Alternative Hypothesis: There is a correlation between the percentage of a country living in a rural area and the GNI of a country. That is, the correlation does not equal 0.

$$H_0: \rho = 0$$

$$H_a: \rho \neq 0$$

### 2. Observed Statistic

```{r}
(obs_stat_corr_rural_logGNI <- cor(health_indicators$Rural, health_indicators$logGNI))
```

### 3. Create null distribution

```{r}
null_dist_corr_rural_logGNI <- rep(NA, 10000)
for (i in 1:10000){
  null_dist_corr_rural_logGNI[i] <-
    cor(sample(health_indicators$Rural), health_indicators$logGNI)
}

# plot the null distribution with a red vertical line for the statistic value
hist(null_dist_corr_rural_logGNI,
     main = "Null Distribution of Correlation between % Population Rural and logGNI",
     xlab = "Correlation",
     xlim = c(-0.8, 0.8),
     nclass = 50)

# So far off that it doesn't even matter
abline(v = obs_stat_corr_rural_logGNI, col = "red")
```

### 4. Calculate p-value

```{r}
(p_value_corr_rural_logGNI <-
   sum(abs(null_dist_corr_rural_logGNI) >= abs(obs_stat_corr_rural_logGNI)) / length(null_dist_corr_rural_logGNI))
```

### 5. Conclusion

TODO: Same here, logGNI or just GNI.

A p-value of 0 is not consistent with there being no correlation between percentage of the population living in rural areas and the GNI. Our null is that the correlation coefficient is 0, so a p-value of 0 means that assuming the null is true, there is a 0% chance that we would observe a correlation coefficient statistic as extreme as ours (-0.7127915).

Thus, we see that GNI and percentage of rural population are quite correlated with each other, and thus can proceed with analyses of either one in our search for a model that predicts economic development (we will use both).

TODO: Should we do a bootstrap confidence interval here, or should we do that somewhere else?

## Linear Model

```{r}
pred_GNI_all <- lm(logGNI ~ Measles + InfMort + LifeExp, data = health_indicators)
summary(pred_GNI_all)
```

It appears that measles and infant mortality are not significant, while LifeExp is. This might be due to a high level of multicollinearity amongst the variables. Specifically, measles likely affects infant mortality, which likely affects life expectancy. We can check this by running a few tests.

```{r}
vif(pred_GNI_all)
```

Since we have a high degree of multicollinearity supported by the variability inflated factor of > 5 for both infant mortality and life expectancy.

```{r}
pred_GNI_LifeExp <- lm(logGNI ~ LifeExp, data = health_indicators)
summary(pred_GNI_LifeExp)
```

In fact, the `R^2` for the model containing all three variables is lower than the `R^2` for the model containing just `LifeExp`.

Let us examine how measles is related to infant mortality, and how infant mortality is related to life expectancy... TODO -> should we pursue this more?

Degree of multicollinearity... in regression model

